---
title: "Exploring causal inference designs"
author: "Francesca Mancini"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(fixest)      # for panel FE models
library(glmmTMB)     # for GLMM comparison models
library(tidyverse)

```

First load the data.

```{r}

all_data <- readRDS("./Data/beewalk_drivers_merged.rds")
str(all_data)
```

We are interested in the effect of temperature on bumblebee abundance. Let's pick one species for now, *Bombus hypnorum*, a common and widespread species.

```{r}

all_data_1sp <- all_data %>%
  filter(latin == "Bombus hypnorum")

```

Now we fit a GLMM with mean temperature as a covariate and a random effect for site. 

```{r}

re_model_temp <- glmmTMB(
  TotalCount ~ temp_mean + (1|GridReference),
  data = all_data_1sp,
  family = nbinom1
)

summary(re_model_temp)

```
Effect of tempearture is significant and the coefficient is negative (-0.03).


Now we fit a fixed effect panel regression with a fixed effect for site. This accounts for time invarying site confounders (variables that vary by site but not by time). 

```{r}

mod_fe <- fenegbin(
  TotalCount ~ temp_mean | GridReference,
  data = all_data_1sp,
  cluster = ~ GridReference # clustered SEs as per Dee et al.
)

summary(mod_fe)



```


The effect of temperature is still negative but smaller and non significant. 
Now I will fit the same model but adding a fixed effect for year. This accounts for any time varying confounders that are equal across sites.

```{r}

mod_fe2 <- fenegbin(
  TotalCount ~ temp_mean | GridReference + Year,
  data = all_data_1sp,
  cluster = ~ GridReference # clustered SEs as per Dee et al.
)

summary(mod_fe2)

```

This changes the sign of the coefficient and the p value is even larger.
It's possible that we don't have enough power due to the fixed effects.

The next two designs allow us to do the same without the inefficiency of the fixed effects.

```{r}

GMCov <- glmmTMB(
  TotalCount ~ temp_mean + site_mean_temp + (1|GridReference),
  data = all_data_1sp,
  family = nbinom1
)

summary(GMCov)


```


This is telling us that the effect of temperature is negative and significant. The contextual effect (the effect of site mean temperature) is close to 0, which in theory means that there are no site level confounders and therefore a mixed model is enough.

But the mean tem[erature and site mean tem[perature are strongly correlated, which makes the previous model not appropriate.


```{r}

ggplot(all_data_1sp,
       aes(x = site_mean_temp, y = temp_mean)) +
  geom_point()

```


The Group Mean Centered design removes this correlation by centering the causal variable around the site average. In this case we create a anomaly temperature variable, by subtracting the site mean temperature scross years from the average temperature for that year.

```{r}

# first derive the site temperature anomaly (temp variable centered around the average site temp)

all_data_1sp <- all_data_1sp %>%
  mutate(temp_cent = temp_mean - site_mean_temp)


GMCent <- glmmTMB(
  TotalCount ~ temp_cent + site_mean_temp + (1|GridReference),
  data = all_data_1sp,
  family = nbinom1
)

summary(GMCent)



```

The effect of mean temperature is negative and the same as estimated by the previous model. But now we have removed the correlation between the two variables. The effect of site mean temperature is also negative, but this includes unmeasured confounders that vary across sites as well so it cannot be interpreted causally.

The differencing designs are useful to separate correlated drivers (confounders that have the same temporal trend as the causal variable) that vary by site as well. We use a second differencing design as we are not particularly interested in site specific trends.

The first step is to calculate the difference in both the response and causal variable between year x and year x-2. Then we fit a linear model.

```{r}

data_sd <- all_data_1sp %>%
  arrange(GridReference, Year) %>%
  group_by(GridReference) %>%
  mutate(
    dy = TotalCount - lag(TotalCount),
    dx = temp_mean - lag(temp_mean),
    d2y = dy - lag(dy),
    d2x = dx - lag(dx)
  ) %>%
  ungroup() %>%
  filter(!is.na(d2y), !is.na(d2x))

sDiff <- lm(d2y ~ d2x, data = data_sd)
summary(sDiff)


```

Effect os still negative and similar in magnitude, but it is not significant.

What does it all mean??


