---
title: "Exploring causal inference designs"
author: "Francesca Mancini"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(fixest)
library(glmmTMB)
library(tidyverse)
library(broom.mixed)   # for tidying glmmTMB models
library(broom)         # for lm models
library(forcats)

```

First load the data.

```{r}

all_data <- readRDS("./Data/beewalk_drivers_merged.rds")
str(all_data)
```

First, I need tofilter the dataset to only keep bumblebee species and impute zeros for all species that were not observed during a transect walk.

```{r}

# filter out all species that are not bumblebees as they are not reliably reported
# filter out all observations not at species level "Bombus sp."
# filter out NAs in the temp columns
# and keep only variables of interest
# finally ungroup dataset
all_data <- all_data %>%
    filter(str_detect(latin, "^Bombus")) %>%
  filter(latin != "Bombus sp.") %>% 
  drop_na(temp_mean, site_mean_temp) %>%
  select(GridReference, section_grid_ref, StartDate, latin, TaxonVersionKey, 
         TotalCount, Year, temp_mean, site_mean_temp) %>%
  ungroup()

# list all species ever observed
all_species <- unique(all_data$TaxonVersionKey)

# identify all sampling events
sampling_events <- all_data %>%
  distinct(GridReference, section_grid_ref, StartDate,
           Year, temp_mean, site_mean_temp,
           .keep_all = FALSE)

# Create a full dataset of all event × species combinations

all_events_expanded <- sampling_events %>%
  crossing(TaxonVersionKey = all_species)



# 5. Prepare species-level data WITHOUT event-level covariates
all_data_species <- all_data %>%
  select(GridReference, section_grid_ref, StartDate,
         Year, TaxonVersionKey, TotalCount)

# 6. Join back counts and species fields 
all_data_zero_filled <- all_events_expanded %>%
  left_join(all_data_species,
            by = c("GridReference", "section_grid_ref",
                   "StartDate", "Year", "TaxonVersionKey")) %>%
  mutate(TotalCount = replace_na(TotalCount, 0))

# create a species-TVK lookup table

sp_tvk_lookup <- unique(all_data[, c("latin", "TaxonVersionKey")])

```


Wewant to fit the following 6 models to each species: 
* GLMM with random effect for site and temperature as covariate  
* Fixed effect model 1 (FE1): fixed effect for site 
* Fixed effect model 2 (FE2): fixed effect for site + year 
* Group Mean Covariate (GMCov) model with a random effect for site, mean site temperature and temperature as covariates 
* Group Mean Centered (GMCent) model with a random effect for site, mean site temperature and temperature centered around the site mean as covariates 
* Second difference model (Diff) 

First we want to create a function tha runs all models for one species.

```{r}
# this catches errors and warnings so we can identify models that did not converge

safe_fit <- function(expr) {
  q <- purrr::quietly(function() expr)()
  tibble(
    model_object = list(q$result),
    warning_msg  = paste(q$warnings, collapse = "; "),
    error_msg    = paste(q$errors, collapse = "; "),
    message_msg  = paste(q$messages, collapse = "; ")
  )
}


fit_all_models_for_species <- function(sp, data) {
  
  dat <- data %>% filter(TaxonVersionKey == sp)
  dat2 <- dat %>% mutate(temp_cent = temp_mean - site_mean_temp)

  # 1) Mixed model (glmmTMB)
  m_re_row <- safe_fit(
    glmmTMB(TotalCount ~ temp_mean + (1|GridReference),
            data = dat, family = nbinom1)
  )
  
  # 2) FE site
  m_fe_row <- safe_fit(
    fenegbin(TotalCount ~ temp_mean | GridReference,
             data = dat, cluster = ~GridReference)
  )
  
  # 3) FE site + year
  m_fe2_row <- safe_fit(
    fenegbin(TotalCount ~ temp_mean | GridReference + Year,
             data = dat, cluster = ~GridReference)
  )
  
  # 4) Group mean covariate
  m_gmcov_row <- safe_fit(
    glmmTMB(TotalCount ~ temp_mean + site_mean_temp + (1|GridReference),
            data = dat, family = nbinom1)
  )
  
  # 5) Group mean centred
  m_gmcent_row <- safe_fit(
    glmmTMB(TotalCount ~ temp_cent + site_mean_temp + (1|GridReference),
            data = dat2, family = nbinom1)
  )
  
  # 6) Second differencing
  dat_sd <- dat %>%
    arrange(GridReference, Year) %>%
    group_by(GridReference) %>%
    mutate(
      dy = TotalCount - lag(TotalCount),
      dx = temp_mean - lag(temp_mean),
      d2y = dy - lag(dy),
      d2x = dx - lag(dx)
    ) %>%
    filter(!is.na(d2y), !is.na(d2x)) %>%
    ungroup()

  m_sdiff_row <- safe_fit(
    lm(d2y ~ d2x, data = dat_sd)
  )

  # Combine all model rows
  model_rows <- bind_rows(
    m_re_row  %>% mutate(model = "RE_glmmTMB"),
    m_fe_row  %>% mutate(model = "FE_site"),
    m_fe2_row %>% mutate(model = "FE_site_year"),
    m_gmcov_row %>% mutate(model = "GM_cov"),
    m_gmcent_row %>% mutate(model = "GM_centered"),
    m_sdiff_row %>% mutate(model = "Second_diff")
  ) %>%
    mutate(species = sp)

  # ---- Extract coefficients if model exists ----
  coef_rows <- model_rows %>%
    mutate(
      tidied = map(model_object, ~ if (is.null(.x)) tibble() else broom::tidy(.x))
    ) %>%
    unnest(tidied)

  return(coef_rows %>%
           select(species, model, term, estimate, std.error, statistic, p.value,
                  warning_msg, error_msg, message_msg))
}

```


Then we apply the function to all species.

```{r}


coef_table <- map_dfr(all_species, fit_all_models_for_species, data = all_data_zero_filled)


```


Filter to only the temperature‑related coefficients.

```{r}

coef_temp <- coef_table %>%
  filter(term %in% c("temp_mean", "temp_cent", "d2x"))

# join species names back in
coef_temp <- left_join(coef_temp, sp_tvk_lookup, by = c("species" = "TaxonVersionKey"))


```


Plot coefficients across species and models.

```{r fig.height=7, fig.width=7}

# we exclude "NHMSYS0000875592", which is B. subterraneus as most models did not converge. 
# This is a very rare species with sparse data

# plot the effects
# but exclude B.distinguendus, B. ruderarius, B.barbutellus and B. ruderatus
# the fixed effect models have very large standard errors
coef_temp_plot <- coef_temp %>%
  filter(!(latin %in% c("Bombus subterraneus",
                        "Bombus distinguendus",
                        "Bombus ruderarius",
                        "Bombus ruderatus",
                        "Bombus barbutellus"))) %>%
  mutate(latin = fct_reorder(latin, estimate, .fun = median, .desc = TRUE))

ggplot(coef_temp_plot,
       aes(x = estimate, y = latin, colour = model)) +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.6, linetype = 2) +
  geom_errorbarh(aes(xmin = estimate - std.error,
                     xmax = estimate + std.error),
                 height = 0, linewidth = 0.7, alpha = 0.9) +
  geom_point(size = 1.8) +
  facet_wrap(~ model, ncol = 3, scales = "fixed") +   
  labs(
    title = "Temperature effect by species",
    x = "Coefficient estimate (± 1 SE)",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_line(colour = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

It looks like there is a range of responses across the species, from negative to positive. The differencing model design always estimates the samllest effects (close to 0) and the effect is not significant for most species. The fixed effects panel regression designs show the largest standard errors, especially the model with both site and year fixed effects. Finally the random effect model, although very precise, it sometimes gets the magnitude and direction of the effect wrong!


Let's make some more comparison plots.
First align the names of the temperature effects across models.

```{r}

# Map each model to its temperature-effect term
temp_term_by_model <- tribble(
  ~model,           ~temp_term,
  "RE_glmmTMB",     "temp_mean",
  "FE_site",        "temp_mean",
  "FE_site_year",   "temp_mean",
  "GM_cov",         "temp_mean",   # within-year temp effect, not the contextual site_mean_temp
  "GM_centered",    "temp_cent",
  "Second_diff",    "d2x"
)

coef_temp_aligned <- coef_temp %>%
  inner_join(temp_term_by_model, by = "model") %>%
  filter(term == temp_term) %>%
  # (optional) keep only species that seem reliable
  filter(!(latin %in% c("Bombus subterraneus",
                        "Bombus distinguendus",
                        "Bombus ruderarius",
                        "Bombus ruderatus",
                        "Bombus barbutellus"))) %>%
  mutate(
    model = factor(model, levels = c("RE_glmmTMB","FE_site","FE_site_year",
                                     "GM_cov","GM_centered","Second_diff")),
    latin = fct_reorder(latin, estimate, .fun = median, .desc = TRUE)
  )

```



First let's compare the direction and significance of the temperature effect across species and models.

```{r, fig.height=9, fig.width=11}

sig_heat <- coef_temp_aligned %>%
  mutate(signif = case_when(
    is.na(p.value)            ~ "NA",
    p.value < 0.05 & estimate > 0 ~ "Pos sig",
    p.value < 0.05 & estimate < 0 ~ "Neg sig",
    TRUE                         ~ "NS"
  ))

ggplot(sig_heat, aes(x = model, y = latin, fill = signif)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("Pos sig" = "#3274A1", "Neg sig" = "#E15759",
                               "NS" = "grey80", "NA" = "grey50")) +
  labs(title = "Sign & significance across species and models",
       x = NULL, y = NULL, fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())

```
This confirms what we saw in the previous plot, the second differencing approach rarely detects a significant result, just like the fixed effect panel regression with fixed effects for both site and year. The fixed effect panel regression with fixed effect for site and the two correlated random effects approaches broadly agree and produce similar results, except for two species B. campestris and B. muscorum. The random effect model detects the most significant effects and sometimes disagrees with other models on the direction of the effect (B. vestalis). 

Then we can do some scatterplots of pairwise comparisons of magnitude and uncertainty of the temperature effect between models across all species. 

```{r fig.height=9, fig.width=11}

library(patchwork)

model_levels <- levels(coef_temp_aligned$model)
pairs <- combn(model_levels, 2, simplify = FALSE)


lims <- range(
  coef_temp_aligned$estimate + coef_temp_aligned$std.error,
  coef_temp_aligned$estimate - coef_temp_aligned$std.error,
  na.rm = TRUE
)



plot_pair <- function(pair) {
  a <- pair[1]
  b <- pair[2]
  
  df <- coef_temp_aligned %>%
    filter(model %in% c(a, b)) %>%
    select(latin, model, estimate, std.error) %>%
    pivot_wider(names_from = model, values_from = c(estimate, std.error), names_sep = "_") %>%
    drop_na()
  
  # rename
  df <- df %>% rename(
    est_a = paste0("estimate_", a),
    est_b = paste0("estimate_", b),
    se_a  = paste0("std.error_", a),
    se_b  = paste0("std.error_", b)
  )
  
  # must return a ggplot object ALWAYS
  if (nrow(df) == 0) {
    return(
      ggplot() +
        annotate("text", 0, 0, label = paste("No data for", a, "vs", b)) +
        theme_void()
    )
  }
  
  ggplot(df, aes(x = est_a, y = est_b)) +
    geom_errorbarh(aes(xmin = est_a - se_a,
                       xmax = est_a + se_a),
                   height = 0, alpha = 0.4) +
    geom_errorbar(aes(ymin = est_b - se_b,
                      ymax = est_b + se_b),
                  width = 0, alpha = 0.4) +
    geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey40") +
    geom_point(size = 1.8, colour = "steelblue") +
    coord_equal() +
    theme_minimal(base_size = 11) +
    labs(title = paste(a, "vs\n", b), x = a, y = b) + 
    coord_equal(xlim = lims, ylim = lims)

}

pair_plots <- map(pairs, plot_pair)

# Correct patchwork usage (no & operator)
final_panel <- wrap_plots(pair_plots) +
  plot_layout(guides = "collect") +
  theme(legend.position = "bottom")

final_panel

```


Now let's compare the width of the standard errors.

```{r}

# Boxplot
ggplot(coef_temp_aligned, aes(x = model, y = std.error, fill = model)) +
  geom_boxplot(outlier.alpha = 0.3, width = 0.7) +
  coord_flip() +
  labs(title = "Uncertainty distribution by model",
       y = "Standard error", x = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

```


As expected, the fixed effect panel regression approaches produce the most uncertain estimates, especially when fixed effects for both site and year are included. The random effect model seems to be the most precise model, but quite comparable with the correlated random effect approaches and the differencing approach.









