---
title: "Predictive performance"
author: "Francesca Mancini"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(yardstick)
library(dplyr)
library(purrr)
library(glmmTMB)
library(fixest)
library(tidyr)
library(ggplot2)

```

We want to compare the predictive performance of the six models fitted in analysis_report.Rmd. This script sets up and runs a cross validation experiment and compares teh performance of different models across species using a set of performance metrics.

First load the data.

```{r}

all_data_zero_filled <- readRDS("./all_data_zero_filled.rds")

```

Define a function to calculate performance metrics. We use three metrics:
* Root Mean Squares Error (RMSE) - this measures the average size of the prediction error, it squares the error so it penalises large mistakes
* Mean Absolute Error (MAE) - measueres average absolute deviation between predictions and true valies, it is easier to interpret as we can say that on average the model is off by X number of inviduals and it is good when abundance distributions are skewed (as is the case with species counts)
* Predictive R-squared (R2) - it compares the model's predictions to a naive baseline model (the mean), if R2 > 0 the model beats the baseline, if R2 < 0 the model performs worse than a mean-only model; it basically compares the explained variance of the model with that of the mean




```{r}


perf_metrics <- function(truth, pred) {
  tibble(
    RMSE = yardstick::rmse_vec(truth, pred),
    MAE  = yardstick::mae_vec(truth, pred),
    R2   = yardstick::rsq_vec(truth, pred)
  )
}


```


Now we define a function to fit the different models.

```{r}

fit_model_set <- function(train_data, temp_center = TRUE) {
  
  # model matrices
  if (temp_center) {
    train_data <- train_data %>%
      mutate(temp_cent = temp_mean - site_mean_temp)
  }
  
  models <- list()
  
  # 1. Mixed model
  models$RE <- glmmTMB(
    TotalCount ~ temp_mean + (1 | GridReference),
    data = train_data, family = nbinom1
  )
  
  # 2. FE site
  models$FE_site <- fenegbin(
    TotalCount ~ temp_mean | GridReference,
    data = train_data, cluster = ~GridReference
  )
  
  # 3. FE site + year
  models$FE_site_year <- fenegbin(
    TotalCount ~ temp_mean | GridReference + Year,
    data = train_data, cluster = ~GridReference
  )
  
  # 4. GM covariate
  models$GM_cov <- glmmTMB(
    TotalCount ~ temp_mean + site_mean_temp + (1 | GridReference),
    data = train_data, family = nbinom1
  )
  
  # 5. GM centered
  models$GM_centered <- glmmTMB(
    TotalCount ~ temp_cent + site_mean_temp + (1 | GridReference),
    data = train_data, family = nbinom1
  )
  
  # 6. Second difference
  train_sd <- train_data %>%
    arrange(GridReference, Year) %>%
    group_by(GridReference) %>%
    mutate(
      dy = TotalCount - lag(TotalCount),
      dx = temp_mean - lag(temp_mean),
      d2y = dy - lag(dy),
      d2x = dx - lag(dx)
    ) %>%
    ungroup() %>%
    filter(!is.na(d2y), !is.na(d2x))
  
  models$Differences <- lm(d2y ~ d2x, data = train_sd)
  
  return(models)
}

```

Now we define a function that generates predictions for each model.

```{r}

predict_model <- function(model, model_name, new_data) {
  
  # -----------------------------
  # 1) RE and GM_cov (glmmTMB)
  # -----------------------------
  if (model_name %in% c("RE", "GM_cov")) {
    return(
      predict(
        model,
        newdata = new_data,
        type = "response",
        re.form = NA
      )
    )
  }
  
  # -----------------------------
  # 2) GM_centered (glmmTMB with centered covariate)
  # -----------------------------
  if (model_name == "GM_centered") {
    
    # MUST recompute centered covariate for test fold
    new_data <- new_data %>%
      mutate(temp_cent = temp_mean - site_mean_temp)
    
    return(
      predict(
        model,
        newdata = new_data,
        type = "response",
        re.form = NA
      )
    )
  }
  
  # -----------------------------
  # 3) fixest FE models
  # -----------------------------
  if (model_name %in% c("FE_site", "FE_site_year")) {
    return(
      predict(
        model,
        newdata = new_data,
        type = "response",
        fixef = FALSE
      )
    )
  }
  
  # -----------------------------
  # 4) Differencing model
  # -----------------------------
  if (model_name == "Differences") {
    
    new_sd <- new_data %>%
      arrange(GridReference, Year) %>%
      group_by(GridReference) %>%
      mutate(
        dy = TotalCount - lag(TotalCount),
        dx = temp_mean - lag(temp_mean),
        d2y = dy - lag(dy),
        d2x = dx - lag(dx)
      ) %>%
      ungroup() %>%
      filter(!is.na(d2x))
    
    pred <- predict(model, newdata = new_sd)
    
    out <- rep(NA, nrow(new_data))
    idx <- match(new_sd$StartDate, new_data$StartDate)
    out[idx] <- pred
    
    return(out)
  }
  
  stop("Unknown model type")
}

```

Now we set up the k-fold cross validation, run the models for each species and then combine the results.

```{r, eval=FALSE}

K <- 5  # number of folds

species_ids <- unique(all_data_zero_filled$TaxonVersionKey)

cv_results <- map_dfr(species_ids, function(sp) {
  
  dat <- all_data_zero_filled %>% filter(TaxonVersionKey == sp)
  
  # assign folds
  set.seed(123)
  dat <- dat %>% mutate(fold = sample(rep(1:K, length.out = n())))
  
  # iterate folds
  fold_results <- map_dfr(1:K, function(f) {
    
    train <- dat %>% filter(fold != f)
    test  <- dat %>% filter(fold == f)
    
    # fit all models
    models <- fit_model_set(train)
    
    # predictions for each model
    preds <- map2_dfc(models, names(models),
      ~ tibble(!!paste0(.y, "_pred") := predict_model(.x, .y, test))
    )
    
    # merge truth + predictions
    df <- bind_cols(
      tibble(
        species = sp,
        fold = f,
        truth = test$TotalCount
      ),
      preds
    )
    
    # long format for metrics
    df_long <- df %>%
      pivot_longer(
        cols = ends_with("_pred"),
        names_to = "model",
        values_to = "pred"
      ) %>%
      mutate(model = sub("_pred$", "", model))
    
    # compute performance metrics per model
    df_long %>%
      group_by(species, model, fold) %>%
      summarise(perf_metrics(truth, pred), .groups = "drop")
    
  })
  
  return(fold_results)
})

```

```{r, echo=FALSE}

cv_results <- readRDS("cv_results.rds")

```


Then summarise performance metrics across folds and species.

```{r}

cv_summary <- cv_results %>%
  group_by(species, model) %>%
  summarise(
    RMSE = mean(RMSE, na.rm = TRUE),
    MAE  = mean(MAE,  na.rm = TRUE),
    R2   = mean(R2,   na.rm = TRUE),
    .groups = "drop"
  )

```

Exclude the species that we know return unreliable models: B. subterraneus, B.distinguendus, B. ruderarius, B.barbutellus and B. ruderatus.


```{r}
sp_tvk_lookup <- readRDS("sp_tvk_lookup.rds")

cv_summary <- left_join(cv_summary, sp_tvk_lookup, by = c("species" = "TaxonVersionKey")) %>%
  filter(!(latin %in% c("Bombus subterraneus",
                        "Bombus distinguendus",
                        "Bombus ruderarius",
                        "Bombus ruderatus",
                        "Bombus barbutellus")))

```

And plot the performance metrics. First RMSE.

```{r}

ggplot(cv_summary, aes(x = model, y = RMSE, colour = model)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme_minimal() +
  labs(title = "Predictive RMSE across models", y = "RMSE", x = NULL) +
  theme(legend.position = "none")

```


Then R2.

```{r}

ggplot(cv_summary, aes(x = model, y = R2, fill = model)) +
  # geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.12) +
  theme_minimal() +
  labs(title = "Predictive R² across models", y = "R² (pred)", x = NULL) +
  theme(legend.position = "none")

```


And finally MAE.

```{r}

ggplot(cv_summary, aes(x = model, y = MAE, colour = model)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme_minimal() +
  labs(title = "Predictive MAE across models", y = "RMSE", x = NULL) +
  theme(legend.position = "none")

```


